---
title: "Practical Machine Learning Course Project"
author: "Dave Richards"
date: "April 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, tidy=TRUE)
```

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Read and clean data

First, read the training data
```{r read1}
training = read.csv("C:/Users/Dave/Documents/R/MachineLearning/pml_project/data/pml-training.csv",stringsAsFactors = FALSE)
dim(training)
str(training)
```

We see a lot of NAs and "" (empty strings). Also, some fields that we would expect to be numeric contain character data, so we investigate one of them further. (I'm omitting the output from this command to save space.)
```{r check_col, include=FALSE}
table(training$kurtosis_roll_dumbbell)
```

We find 5 "DIV/0!" values, so let's reread the data, changing them as well as empty strings and "NA" strings to NA.
```{r read_train}
training = read.csv("C:/Users/Dave/Documents/R/MachineLearning/pml_project/data/pml-training.csv",stringsAsFactors = FALSE,na.strings = c("NA", "", "#DIV/0!"))
sum(training == '#DIV/0', na.rm = TRUE)
sum(training == '', na.rm = TRUE)
```

Create a new dataset with only the columns that are not more than 95% NA.
```{r train_good}
keep_train <- apply(training, 2, function(x) mean(is.na(x)) < 0.95)
train_good <- training[,keep_train]
head(train_good)
```

The first seven columns are metadata so let's remove them as well.
```{r remove_meta}
train_good <- train_good[,-(1:7)]
dim(train_good)
sum(complete.cases(train_good))
```

We read and pre-process the testing data the same way.
```{r read_test}
testing = read.csv("~/R/MachineLearning/pml_project/data/pml-testing.csv", stringsAsFactors = FALSE, na.strings=c("NA", "", "#DIV/0!"))
test_good <- testing[,keep_train]
test_good <- test_good[,-(1:7)]
dim(test_good)
sum(complete.cases(test_good))
```

## Create and train model

We are going to use the Random Forest model to predict the class for the test cases. Random Forest is very accurate but can take a long time to run. We will improve performance by using parallel processing. **Thanks to Len Greski** for his overviews on setting up parallel processing as well as configuring Git.

Set seed for reproducibility.
```{r seed}
set.seed(32464)
```

Include caret package and configure parallel processing. Caret automates k-fold cross-validation and estimates out of sample error.
```{r parallel}
library(caret)
library(parallel)
library(doParallel)
detectCores()
cluster <- makeCluster(detectCores() - 1)  # Leave one core for the OS
registerDoParallel(cluster)
```

Configure the trainControl object for 10-fold cross-validation.
```{r trainControl}
fitControl <- trainControl(method = "cv", number = 10, allowParallel = TRUE)
```

Develop the training model using random forests.
```{r model, cache=TRUE}
x <- train_good[,-53]
y <- train_good[,53]
fit <- train(x, y, method="rf", data=train_good, trControl=fitControl)  # This can take a long time to run.
```

De-register the parallel processing cluster
```{r stop_cluster}
stopCluster(cluster)
```

We check accuracy (>99%) and confusion matrix of model.
```{r mod_check}
fit
confusionMatrix.train(fit)
```

Now we use the model to predict on test data.
```{r predict}
x <- test_good[,-53]
prediction <- predict(fit, x)
prediction
```

## Conclusion

The model predicted all 20 of the observations in the testing dataset correctly.
